{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-08-24 ident_manu.Rnw
methods should reflect p1z1, added obser uncert info

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.  The sensitivity of \ac{do} to perturbations of all relevant parameters for the 108 equations was estimated from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006 by running FishTank at a timestep of five minutes.  Initial screening removed 73 irrelevant parameters that were removed for several reasons, e.g., a parameter was only relevant for the hydrodynamic model, was considered a physical constant, or had no effect given initial conditions.

<<echo = FALSE>>=
# remove irrelevant parameters, from 'ignore/noparams.csv'
rmpars <- c('Kw_1', 'Kcdom_1', 'Kspm_1', 'Kchla_1', 'aw490_1', 'astarOMR_1', 'astarOMBC_1', 'KTg1(nospA+nospZ)_1', 'KTg1(nospA+nospZ)_2', 'KTg1(nospA+nospZ)_3', 'KTg1(nospA+nospZ)_4', 'KTg1(nospA+nospZ)_5', 'KTg1(nospA+nospZ)_6', 'KTg1(nospA+nospZ)_7', 'KTg1(nospA+nospZ)_8', 'KTg2(nospA+nospZ)_1', 'KTg2(nospA+nospZ)_2', 'KTg2(nospA+nospZ)_3', 'KTg2(nospA+nospZ)_4', 'KTg2(nospA+nospZ)_5', 'KTg2(nospA+nospZ)_6', 'KTg2(nospA+nospZ)_7', 'KTg2(nospA+nospZ)_8', 'Ea_R(nospA+nospZ)_1', 'Ea_R(nospA+nospZ)_2', 'Ea_R(nospA+nospZ)_3', 'Ea_R(nospA+nospZ)_4', 'Ea_R(nospA+nospZ)_5', 'Ea_R(nospA+nospZ)_6', 'Ea_R(nospA+nospZ)_7', 'Ea_R(nospA+nospZ)_8', 'beta_1', 'beta_2', 'beta_3', 'beta_4', 'beta_5', 'beta_6', 'QmaxN_1', 'QmaxN_2', 'QmaxN_3', 'QmaxN_4', 'QmaxN_5', 'QmaxN_6', 'QmaxP_1', 'QmaxP_2', 'QmaxP_3', 'QmaxP_4', 'QmaxP_5', 'QmaxP_6', 'nfQs_1', 'nfQs_2', 'nfQs_3', 'nfQs_4', 'nfQs_5', 'nfQs_6', 'KG1_R_1', 'KG2_R_1', 'KG1_BC_1', 'KG2_BC_1', 'pCO2_1', 'stoich_x1R_1', 'stoich_y1R_1', 'stoich_x2R_1', 'stoich_y2R_1', 'stoich_x1BC_1', 'stoich_y1BC_1', 'stoich_x2BC_1', 'stoich_y2BC_1', 'sink OM1_R_1', 'sink OM2_R_1', 'sink OM1_BC_1', 'sink OM2_BC_1', 'CF_SPM_1')

# number of parameters in each category
nparms <- parcats2(as_df = TRUE) %>% 
  filter(!shrt %in% rmpars) %>% 
  .$cats %>% 
  table %>% 
  as.list
@

As noted above, FishTank includes six phytoplankton and two zooplankton groups to characterize greater complexity in community structure and foodweb dynamics.  However, structural equations for each group do not vary considerably such that variation in parameter values primarily control differences between the groups, e.g., large-bodied or small-bodied plankton, slow-growing or fast-growing plankton, etc..  Initial analyses indicated that parameter sensitivity of dissolved oxygen was identical within the six phytoplankton and zooplankton groups.  To remove obvious redundancies in the model, the sensitivity analyes were conducted using only one phytoplankton and one zooplankton group.  This further reduced the parameter set....  For simplicity, the remaining parameters were grouped in  to one of six categories based on their respective equations: optics ($n = $ \Sexpr{nparms$Optics} parameters), organic matter (\Sexpr{nparms$`Organic Matter`}), phytoplankton (\Sexpr{nparms$Phytoplankton}), temperature (\Sexpr{nparms$Temperature}), and zooplankton (\Sexpr{nparms$Zooplankton}).  A full description is available as an appendix in \citepalias{LehrterIR}.

@@

The analysis focused on sensitivity of \ac{do} in the \ac{zerod} FishTank model to identify parameters that may affect spatial and temporal variation of hypoxia in the larger model.  A local sensitivity analysis was performed for each of the 178 parameters using a simple perturbation approach to evaluate the change in \ac{do} from the original parameter values.  The analyses relied exlusively on concepts used in the FME package developed for the R statistical programming language \citep{Soetaert10}. Each parameter was perturbed by 50\% of its original value and the model was executed to obtain an estimate of the effect on \ac{do}.  For each perturbation, a sensitivity value $S$ was estimated for each time step $i$ given a set value for parameter $j$ as:

@@

\noindent where the estimate depended on the change in the predicted value for response variable $y$ divided by the change in the parameter $\Theta_j$ multiplied by the quotient of scaling factors $w$ for each.  The scaling factors, $w_{\Theta_j}$ for the parameter $\Theta_j$ and $w_{y_i}$ for response variable $y_i$, were set as the default value of the unperturbed parameter and the initial value of $y$ at time $= 0$ \citep{Soetaert10}.  The scaling ensures the estimates are unitless such that the relative magnitudes provide a comparison for model sensitivity to parameter changes and state variables that differ in scale.  Estimates for $S_{ij}$ were summarized as $L1$ and $L2$ across the time series to obtain individual sensitivity values of \ac{do} in response to a change in parame ter $j$:

@@

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 178 evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter.  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability, and the difficulty of choosing from many combinations.

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed with the assumption that parameters will be selected with preference for those with high sensitivity and identifability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within parameter category,
\item independent of parameter category,
\item or considering all categories equally
\end{inparaenum}.  The selection rules produced seven subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

\subsection{Identifiability and structural uncertainty}

As a final analysis, the effects of model structure on parameter identifiability were evaluated for \ac{do} predictions.  The FishTank model includes several processes that can be included or excluded from the equations based on expected conditions or available data.  These `switches' are conceptually different from model parameters as they define the use of explicit equations or processes that affect the model structure.  As such, a comparison of sensitivity and identifiability given different equations uses parameter uncertainty as a means to begin addressing componenents of structural uncertainty.  Switches in FishTank include different structural equations for the vertical attenuation of light through the water column (inherent or apparent optical properties, \citealt{Penta09,Eldridge10}) and chlorophyll to carbon ratio models (fixed or dynamic given light and nutrients, \citealt{Cloern95}).  Several switches also affect phytoplankton growth including different models for specific growth and effects of temperature, light dependence, nutrient uptake, and internal cell quotas \citepalias[][references therein]{LehrterIR}.  Parameter identifiability was evaluated using three scenarios of simple, intermediate, and complex combinations of model switches.

@@

<<sensalltile, eval = T, echo = F, fig.height = 7, fig.width = 6, out.width = '\\textwidth', fig.cap = 'Sensitivity values (L1, \\cref{l1}) for local analyses of all state variables.  Values are in log-space for visual comparison. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.'>>=
data(sens_ests_cat_all)

>>>>

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.  The sensitivity of \ac{do} to perturbations of all relevant parameters for the 108 equations was estimated from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006 by running FishTank at a timestep of five minutes.

<<echo = FALSE>>=
pars <- parcats2(as_df = TRUE)$shrt %>% 
  as.character
rmpars <- c('Kw_1', 'Kcdom_1', 'Kspm_1', 'Kchla_1', 'aw490_1', 'astarOMR_1', 'astarOMBC_1', 'KTg1(nospA+nospZ)_1', 'KTg1(nospA+nospZ)_2', 'KTg1(nospA+nospZ)_3', 'KTg1(nospA+nospZ)_4', 'KTg1(nospA+nospZ)_5', 'KTg1(nospA+nospZ)_6', 'KTg1(nospA+nospZ)_7', 'KTg1(nospA+nospZ)_8', 'KTg2(nospA+nospZ)_1', 'KTg2(nospA+nospZ)_2', 'KTg2(nospA+nospZ)_3', 'KTg2(nospA+nospZ)_4', 'KTg2(nospA+nospZ)_5', 'KTg2(nospA+nospZ)_6', 'KTg2(nospA+nospZ)_7', 'KTg2(nospA+nospZ)_8', 'Ea_R(nospA+nospZ)_1', 'Ea_R(nospA+nospZ)_2', 'Ea_R(nospA+nospZ)_3', 'Ea_R(nospA+nospZ)_4', 'Ea_R(nospA+nospZ)_5', 'Ea_R(nospA+nospZ)_6', 'Ea_R(nospA+nospZ)_7', 'Ea_R(nospA+nospZ)_8', 'beta_1', 'beta_2', 'beta_3', 'beta_4', 'beta_5', 'beta_6', 'QmaxN_1', 'QmaxN_2', 'QmaxN_3', 'QmaxN_4', 'QmaxN_5', 'QmaxN_6', 'QmaxP_1', 'QmaxP_2', 'QmaxP_3', 'QmaxP_4', 'QmaxP_5', 'QmaxP_6', 'nfQs_1', 'nfQs_2', 'nfQs_3', 'nfQs_4', 'nfQs_5', 'nfQs_6', 'KG1_R_1', 'KG2_R_1', 'KG1_BC_1', 'KG2_BC_1', 'pCO2_1', 'stoich_x1R_1', 'stoich_y1R_1', 'stoich_x2R_1', 'stoich_y2R_1', 'stoich_x1BC_1', 'stoich_y1BC_1', 'stoich_x2BC_1', 'stoich_y2BC_1', 'sink OM1_R_1', 'sink OM2_R_1', 'sink OM1_BC_1', 'sink OM2_BC_1', 'CF_SPM_1')
rmpars <- c(rmpars, 'respb_7') # there's an extra of this one
pars <- pars[!pars %in% rmpars]
p1z1pars <- grep('_1$|_7$', pars, value = TRUE) # get parameters for one phyto, one zoop (_7 is for temp)

# number of parameters in each category
nparms <- parcats2(as_df = TRUE) %>% 
  filter(!shrt %in% rmpars) %>% 
  filter(grepl('_1$|_7$', shrt)) %>% 
  .$cats %>% 
  table %>% 
  as.list

Irrelevant parameters were removed for several reasons from the analysis.  Parameters that were removed were not relevant for the \ac{zerod} model (i.e., hydrodynamic parameters), were considered physical constants, or had no effect given initial conditions.  Additionally, FishTank includes six phytoplankton and two zooplankton groups to describe complexity in community structure and foodweb dynamics.  However, structural equations for each group do not vary considerably such that variation in parameter values primarily controls differences between the groups, e.g., large-bodied or small-bodied plankton, slow-growing or fast-growing plankton, etc.  Initial analyses indicated that parameter sensitivity of dissolved oxygen was identical within the six phytoplankton and zooplankton groups.  To remove obvious redundancies in the model, the sensitivity analyses were conducted using only one phytoplankton and one zooplankton group.  The final parameter set that was evaluated included \Sexpr{length(p1z1pars)} parameters that were further grouped into one of six categories based on applicable biogeochemical components of the model: optics ($n = $ \Sexpr{nparms$Optics} parameters), organic matter (\Sexpr{nparms$`Organic Matter`}), phytoplankton (\Sexpr{nparms$Phytoplankton}), temperature (\Sexpr{nparms$Temperature}), and zooplankton (\Sexpr{nparms$Zooplankton}).  A full description of the model parameters is available as an appendix in \citetalias{LehrterIR}.

@@

The analysis focused on sensitivity of \ac{do} in the \ac{zerod} FishTank model to identify parameters that may affect spatial and temporal variation of hypoxia in the larger model.  A local sensitivity analysis was performed for each of the parameters using a simple perturbation approach to evaluate the change in \ac{do} from the original parameter values.  The analyses relied exlusively on concepts used in the FME package developed for the R statistical programming language \citep{Soetaert10}. Each parameter was perturbed by 50\% of its original value and the model was executed to obtain an estimate of the effect on \ac{do}.  For each perturbation, a sensitivity value $S$ was estimated for each time step $i$ given a set value for parameter $j$ as:

@@

\noindent where the estimate depended on the change in the predicted value for response variable $y$ divided by the change in the parameter $\Theta_j$ multiplied by the quotient of scaling factors $w$ for each.  The scaling factors, $w_{\Theta_j}$ for the parameter $\Theta_j$ and $w_{y_i}$ for response variable $y_i$, were set as the default value of the unperturbed parameter and the predicted value of $y_i$ after perturbation \citep{Soetaert10}.  The scaling ensures the estimates are unitless such that the relative magnitudes allows comparisons of model sensitivity to parameter changes and state variables that differ in scale.  Estimates for $S_{ij}$ were summarized as $L1$ and $L2$ across the time series to obtain individual sensitivity values of \ac{do} in response to a change in parame ter $j$:

@@

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the \Sexpr{length(p1z1pars)} evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter.  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability, and the difficulty of choosing from many combinations.

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed with the assumption that parameters will be selected with preference for those with high sensitivity and identifability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences by model users given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within parameter category,
\item independent of parameter category,
\item or considering all categories equally.
\end{inparaenum} The selection rules produced eight subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

\subsection{Observational and structural uncertainty}

As a final analysis, the effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability to infer the effects of observational and structural uncertainty on model output to create a more comprehensive evaluation.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on which nutrient is limiting .  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values that were estimated from the observed data.

The effects of model structure on parameter identifiability were evaluated by including or excluding specific components of the model.  The FishTank model includes several processes that are based on expected conditions or available data.  These `switches' are conceptually different from model parameters as they define the use of explicit equations or processes included in the model structure.  As such, a comparison of sensitivity and identifiability given different equations uses parameter uncertainty to begin addressing componenents of structural uncertainty.  Switches in FishTank include different structural equations for the vertical attenuation of light through the water column (inherent or apparent optical properties, \citealt{Penta09,Eldridge10}) and chlorophyll to carbon ratio models (fixed or dynamic given light and nutrients, \citealt{Cloern95}).  Several switches also affect phytoplankton growth including different models for specific growth and effects of temperature, light dependence, nutrient uptake, and internal cell quotas \citepalias[][references therein]{LehrterIR}.  Parameter identifiability was evaluated using three scenarios of simple, intermediate, and complex combinations of model switches.

@@

<<sensalltile, eval = T, echo = F, fig.height = 7, fig.width = 6, out.width = '\\textwidth', fig.cap = 'Sensitivity values (L1, \\cref{l1}) for local analyses of all state variables. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.'>>=
data(sens_ests_cat_all)

### 2016-08-24 ident_manu.tex
methods should reflect p1z1, added obser uncert info

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Mon Aug 22 16:13:24 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.  The sensitivity of \ac{do} to perturbations of all relevant parameters for the 108 equations was estimated from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006 by running FishTank at a timestep of five minutes.  Initial screening removed 73 irrelevant parameters that were removed for several reasons, e.g., a parameter was only relevant for the hydrodynamic model, was considered a physical constant, or had no effect given initial conditions.

As noted above, FishTank includes six phytoplankton and two zooplankton groups to characterize greater complexity in community structure and foodweb dynamics.  However, structural equations for each group do not vary considerably such that variation in parameter values primarily control differences between the groups, e.g., large-bodied or small-bodied plankton, slow-growing or fast-growing plankton, etc..  Initial analyses indicated that parameter sensitivity of dissolved oxygen was identical within the six phytoplankton and zooplankton groups.  To remove obvious redundancies in the model, the sensitivity analyes were conducted using only one phytoplankton and one zooplankton group.  This further reduced the parameter set....  For simplicity, the remaining parameters were grouped in  to one of six categories based on their respective equations: optics ($n = $ 4 parameters), organic matter (12), phytoplankton (132), temperature (8), and zooplankton (22).  A full description is available as an appendix in \citepalias{LehrterIR}.

@@

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 178 evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter.  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability, and the difficulty of choosing from many combinations.

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed with the assumption that parameters will be selected with preference for those with high sensitivity and identifability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within parameter category,
\item independent of parameter category,
\item or considering all categories equally
\end{inparaenum}.  The selection rules produced seven subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

\subsection{Identifiability and structural uncertainty}

As a final analysis, the effects of model structure on parameter identifiability were evaluated for \ac{do} predictions.  The FishTank model includes several processes that can be included or excluded from the equations based on expected conditions or available data.  These `switches' are conceptually different from model parameters as they define the use of explicit equations or processes that affect the model structure.  As such, a comparison of sensitivity and identifiability given different equations uses parameter uncertainty as a means to begin addressing componenents of structural uncertainty.  Switches in FishTank include different structural equations for the vertical attenuation of light through the water column (inherent or apparent optical properties, \citealt{Penta09,Eldridge10}) and chlorophyll to carbon ratio models (fixed or dynamic given light and nutrients, \citealt{Cloern95}).  Several switches also affect phytoplankton growth including different models for specific growth and effects of temperature, light dependence, nutrient uptake, and internal cell quotas \citepalias[][references therein]{LehrterIR}.  Parameter identifiability was evaluated using three scenarios of simple, intermediate, and complex combinations of model switches.

@@

\caption[Sensitivity values (L1, \cref{l1}) for local analyses of all state variables]{Sensitivity values (L1, \cref{l1}) for local analyses of all state variables.  Values are in log-space for visual comparison. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.}\label{fig:sensalltile}
\end{figure}

>>>>

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Tue Aug 23 17:18:33 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.  The sensitivity of \ac{do} to perturbations of all relevant parameters for the 108 equations was estimated from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006 by running FishTank at a timestep of five minutes.

Irrelevant parameters were removed for several reasons from the analysis.  Parameters that were removed were not relevant for the \ac{zerod} model (i.e., hydrodynamic parameters), were considered physical constants, or had no effect given initial conditions.  Additionally, FishTank includes six phytoplankton and two zooplankton groups to describe complexity in community structure and foodweb dynamics.  However, structural equations for each group do not vary considerably such that variation in parameter values primarily controls differences between the groups, e.g., large-bodied or small-bodied plankton, slow-growing or fast-growing plankton, etc.  Initial analyses indicated that parameter sensitivity of dissolved oxygen was identical within the six phytoplankton and zooplankton groups.  To remove obvious redundancies in the model, the sensitivity analyses were conducted using only one phytoplankton and one zooplankton group.  The final parameter set that was evaluated included 51 parameters that were further grouped into one of six categories based on applicable biogeochemical components of the model: optics ($n = $ 4 parameters), organic matter (12), phytoplankton (22), temperature (2), and zooplankton (11).  A full description of the model parameters is available as an appendix in \citetalias{LehrterIR}.

@@

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 51 evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter.  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability, and the difficulty of choosing from many combinations.

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed with the assumption that parameters will be selected with preference for those with high sensitivity and identifability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences by model users given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within parameter category,
\item independent of parameter category,
\item or considering all categories equally.
\end{inparaenum} The selection rules produced eight subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

\subsection{Observational and structural uncertainty}

As a final analysis, the effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability to infer the effects of observational and structural uncertainty on model output to create a more comprehensive evaluation.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on the limiting nutrient .  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values estimated from the observed data.

The effects of model structure on parameter identifiability were evaluated by including or excluding specific components of the model.  The FishTank model includes several processes that are based on expected conditions or available data.  These `switches' are conceptually different from model parameters as they define the use of explicit equations or processes included in the model structure.  As such, a comparison of sensitivity and identifiability given different equations uses parameter uncertainty to begin addressing componenents of structural uncertainty.  Switches in FishTank include different structural equations for the vertical attenuation of light through the water column (inherent or apparent optical properties, \citealt{Penta09,Eldridge10}) and chlorophyll to carbon ratio models (fixed or dynamic given light and nutrients, \citealt{Cloern95}).  Several switches also affect phytoplankton growth including different models for specific growth and effects of temperature, light dependence, nutrient uptake, and internal cell quotas \citepalias[][references therein]{LehrterIR}.  Parameter identifiability was evaluated using three scenarios of simple, intermediate, and complex combinations of model switches.

@@

\caption[Sensitivity values (L1, \cref{l1}) for local analyses of all state variables]{Sensitivity values (L1, \cref{l1}) for local analyses of all state variables. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.}\label{fig:sensalltile}
\end{figure}
